---
title: "Data Analysis Exercise"
---

```{r}

# load dslabs package
library("dslabs") 
library("dplyr")
library("ggplot2")

# get an overview of data structure
str(gapminder)

# get a summary of data
summary(gapminder)

# determine the type of object gapminder is
class(gapminder)

africadata <- gapminder %>%
  filter(continent == "Africa")

# Create a new data frame with only 'infant_mortality' and 'life_expectancy'
africa_mort_life <- africadata %>%
  select(infant_mortality, life_expectancy)

# Create another data frame with only 'population' and 'life_expectancy'
africa_pop_life <- africadata %>%
  select(population, life_expectancy)

# Check the structure of the first new object
str(africa_mort_life)
# Check a summary of the data to understand distributions and NA presence
summary(africa_mort_life)

# Check the structure of the second new object
str(africa_pop_life)
# Summary of population and life expectancy
summary(africa_pop_life)
```

I created two new data frames from the "africadata" object:  
- "africa_mort_life": Contains only the "infant_mortality" and "life_expectancy" columns.
- "africa_pop_life": Contains only the "population" and "life_expectancy" columns.

Both data frames contain 2907 rows (one for each African country-year observation) and 2 columns. I used str() to inspect the structure (data types and column names) and summary() to examine the distribution and detect missing values.

```{r}
# Load dplyr package
library(dplyr)

# Create the object with infant_mortality and life_expectancy
africa_mort_life <- africadata %>%
  select(infant_mortality, life_expectancy)

# Plotting  Life expectancy vs infant mortality 
ggplot(africa_mort_life, aes(x = infant_mortality, y = life_expectancy)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  labs(
    title = "Life expectancy vs Infant mortality in Africa",
    x = "Infant Mortality",
    y = "Life Expectancy (years)"
  ) +
  theme_minimal()

```

This plot shows a  negative correlation between infant mortality and life expectancy. Countries with higher infant mortality tend to have lower life expectancy. 

```{r}
# Create population vs life expectancy data from africadata
africa_pop_life <- africadata %>%
  select(population, life_expectancy)

# Plotting Life expectancy vs pop size 
ggplot(africa_pop_life, aes(x = population, y = life_expectancy)) +
  geom_point(alpha = 0.6, color = "darkgreen") +
  scale_x_log10() +  # log scale for population
  labs(
    title = "Life expectancy vs Population size",
    x = "Population (log scale)",
    y = "Life Expectancy (years)"
  ) +
  theme_minimal()

```

Here, we observe a positive correlation between population size and life expectancy.

However, the data displays streaks of points. When we examine the original africadata again, we notice that each country has multiple entries over time (from 1960 to 2016). These streaks represent temporal data points for each country, where each streak corresponds to a single country progressing over time.

```{r}
# Now perform analysis for year 20000
africadata_year2000 <- subset(africadata, year == 2000)
# get an overview of data structure
str(africadata_year2000)

 # summary of data
summary(africadata_year2000)

ggplot(africadata_year2000, aes(x = infant_mortality, y = life_expectancy)) +
  geom_point(color = "steelblue", alpha = 0.7) +
  labs(
    title = "Life Expectancy vs Infant Mortality -year 2000",
    x = "Infant mortality",
    y = "Life expectancy (years)"
  ) +
  theme_minimal()

```

```{r}

# Now we will apply models to our analysis of year 2000 to see if they are significant correlations:

# Model 1: Predict life expectancy using infant mortality
fit1 <- lm(life_expectancy ~ infant_mortality, data = africadata_year2000)

# Model 2: Predict life expectancy using population size
fit2 <- lm(life_expectancy ~ population, data = africadata_year2000)

# Print model summaries to screen
summary(fit1)

lm(formula = life_expectancy ~ infant_mortality, data = africadata_year2000)

summary(fit2)

lm(formula = life_expectancy ~ population, data = africadata_year2000)

```

I fitted two simple linear regression models using the lm() function on African countries in the year 2000:

Model 1 (fit1): Life expectancy as a function of infant mortality.
- The model shows a negative correlation between infant mortality and life expectancy.
- The estimate for infant mortality is negative and statistically significant (very low p-value), which confirms this inverse relationship.

Model 2 (fit2): Life expectancy as a function of population size.
- This model shows not a strong correlation between population size and life expectancy.
- The p-value and r-quared are not significant.

In the year 2000, infant mortality is a stronger predictor of life expectancy than population size, at least in the year 2000 in Africa. 


```{r}
# Additional Data Exploration – this part was contributed by Talia C. Swanson

library(dslabs)
library(dplyr)
library(ggplot2)

# Upload chosen dataset
data("historic_co2")

# Displaying all points in the historic_co2 dataset
historic_co2

# Check structure of data
str(historic_co2)

```

The str( ) function shows the structure of the historic_co2 dataset. It has 694 total observations and 3 variables, 
  year – the year of measurement
  co2 – the CO₂ concentration 
  source – the measurement source 

This helps us understand the type of data we’re working with and confirms that the dataset is ready for analysis and visualization.

```{r}
# You will preview beginning data and look at your descriptive stats
head(historic_co2)
summary(historic_co2)
```

The head( ) function displays the first 6 rows of the dataset. From this, we can see that CO₂ concentrations start around 316 ppm in 1959 from Mauna Loa and that measurements are chronological.

The summary( ) function provides descriptive statistics for each variable we have, year, source, and co2 concentration. 


```{r}
# Plotting CO2 concentration over all time
ggplot(historic_co2, aes(x = year, y = co2)) +
  geom_line(color = "blue", size = 1) +
  labs(title = "Historical CO2 Concentration Over Time",
       x = "Year",
       y = "CO2 (ppm)") +
  theme_minimal()
```

The historic_co2 dataset contains atmospheric CO₂ concentrations measured over time with the first table showing the first few observations within the dataset. Within this visualization, CO₂ concentrations have increased steadily from the mid 1900's onward. There are small fluctuations of increasing and decreasing behavior throughout the years, however, the general trend is moving upwards indictaing a long-term rise in atmospheric CO₂ levels. 


```{r}
# Filter for recent years 2000 onwards as target
recent_co2 <- historic_co2 %>% filter(year >= 2000)

# Display the table of points used in the plot (for recent years)
recent_co2

# This will plot our graph of atmospheric CO2 for years 2000 and above
ggplot(recent_co2, aes(x = year, y = co2)) +
  geom_line(color = "red") +
  geom_point(color = "darkred") +
  labs(title = "CO2 Levels Since 2000",
       x = "Year",
       y = "CO2 (ppm)") +
  theme_minimal()
```

This visualization focuses on CO₂ levels from the year 2000 to the present. During this period, CO₂ concetrations continue to increase at a steady rate without any major decreases or fluctuations. This suggests that atmospheric CO₂ have consistently risen in more recent years.